---
title: "Assignment 4"
format: html
editor: visual
---
_Kelly Primrose in collaboration with Diana Schroder_

#Library Addition & Data Importing
```{r}
#| echo:false

library(tidyverse)
library(tinytex)
library(ggplot2)
library(lubridate)
library(sf)
library(tigris)

cabi <- read.csv("202301-capitalbikeshare-tripdata.csv")

```

We decided to use the Capital Bikeshare ridership data from January 2023 to examine the paterns of CaBi riders at various points in time. 

*Graph 1: Origin Points of Rides in DC*
```{r}
start_station <- st_as_sf(cabi, coords = c("start_lng", "start_lat"))%>% 
  st_set_crs(value = 4326)

dc <- states(progress_bar = FALSE) %>%
 filter(STUSPS == "DC") 

ggplot() +
 geom_sf(data = dc) +
 geom_sf(data = start_station,
         aes(color = member_casual)) +
 theme_void()+
  labs(title = "Origin Points of Rides in DC", 
       subtitle = "Majority of ridership is by members",
       caption = "Data source: Capitol Bikeshare Trip History Data",
       color = "Capitol Bikeshare Membership Status")

ggsave(filename = "GG1_DCOrigin.png")
```


*Graph 2: Ridership Across DMV by Day of Week*
```{r}
# Put end time variable into date format and extract day of the week
end_time <- ymd_hms(cabi$ended_at)
day <- wday(end_time)
cabi %>% 
  ggplot() +
  geom_bar(mapping = aes(x = day))+
  labs(title = "Ridership By Day of Week", 
       caption = "Data publicly available on the Capital Bikeshare website.")+
  xlab("Day of week")+
  ylab("Number of rides taken")

ggsave(filename = "GG2_DayBar.png")
```

*Graph 3: Top 10 Most Popular Destination Stations
```{r}

#replacing empty instances of end_station_name with "Abandoned bike", as these unreturned bikes were likely left somewhere. 

cabi <- cabi%>%
  mutate(end_station_name = ifelse(end_station_name == "", "Abandoned bike", end_station_name))

#grouping by end station and counting in descending order to see which stations are most popular destinations
cabi%>%
  group_by(end_station_name)%>%
  summarize(count = n())%>%
  arrange(desc(count))%>%
  filter(count > 1630)%>%
  ggplot() +
  geom_count(mapping = aes(x=end_station_name, y = count))+
  theme(axis.text.x = element_text(angle = 60, hjust = 1))

ggsave(filename = "populardests.png")

```

*Graph 4: Trip Duration by Members Versus Casual Riders*

```{r}
end_time <- ymd_hms(cabi$ended_at)
start_time <- ymd_hms(cabi$started_at)

cabiduration <- data.frame(start_time, end_time)

cabiduration%>%
  difftime(end_time, start_time, units = "minutes")

ended_at <- as.POSIXct.Date("ended_at")
  mutate(duration = difftime(started_at, ended_at, units = "minutes"))

```

